---
title: "MLB Day Night Game Analysis"
author: "cwhaley"
date: "`r Sys.Date()`"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background

This analysis will look at MLB team performances, comparing day games and night games. We'll
use the following packages for MLB game data, project workflow, and data manipulation:

```{r load packages}
library(baseballr)  # game/season data
library(tidyverse)  # data manipulation
library(here)       # project workflow, easy folder navigation

data(teams_lu_table)  # reference data for teams (divisions)
```

## Clean Team and Division Names

Here we will shorten the division names by assigning an acronym to the league -- American
League becomes "AL", and National League becomes "NL". The reference table also has the wrong name
for Cleveland (i.e., Indians) so we will change this to the current name (i.e., Guardians).

```{r division and team names, echo=FALSE}
# clean up division names
teams_divisions <- teams_lu_table |> 
  filter(sport.name == 'Major League Baseball') |> 
  select(name, league.name, division.name) |> 
  mutate(division = case_when(
    league.name == 'American League' ~ str_replace_all(division.name, 'American League', 'AL'),
    league.name == 'National League' ~ str_replace_all(division.name, 'National League', 'NL')
  )) |> 
  rename(league = league.name) |> 
  select(-division.name)

# fix the Cleveland Guardians team name in the lookup division table
teams_divisions$name <- ifelse(teams_divisions$name == 'Cleveland Indians',
                          'Cleveland Guardians',
                          teams_divisions$name)
```



```{r pivot data, echo=FALSE}
# basically pivot the wide format of the data to long
away_teams <- mlb_schedule(season = 2023, level_ids = "1") |> 
  filter(series_description == 'Regular Season' & status_detailed_state == 'Final') |> 
  select(date, game_pk, day_night, teams_away_team_name, teams_away_is_winner) |> 
  mutate(teams_away_is_winner = case_when(
    teams_away_is_winner == TRUE ~ 1,
    TRUE ~ 0
  )) |> 
  rename(team = teams_away_team_name, win_indicator = teams_away_is_winner)

home_teams <-  mlb_schedule(season = 2023, level_ids = "1") |> 
  filter(series_description == 'Regular Season' & status_detailed_state == 'Final') |> 
  select(date, game_pk, day_night, teams_home_team_name, teams_home_is_winner) |> 
  mutate(teams_home_is_winner = case_when(
    teams_home_is_winner == TRUE ~ 1,
    TRUE ~ 0
  )) |> 
  rename(team = teams_home_team_name, win_indicator = teams_home_is_winner)

# union back into table, long format
all_teams <- bind_rows(away_teams, home_teams)

# add division names
all_teams <- all_teams |> 
  inner_join(teams_divisions, by = c("team"="name"))
```

```{r aggregate wins and percentages, echo=FALSE}
# wins per team per day-night
all_teams_agg <- all_teams |> 
  group_by(day_night, team) |> 
  summarise(wins = sum(win_indicator),
            num_games = n_distinct(game_pk),
            day_night_win_pct = wins/num_games) |> 
  arrange(team)

# add in division info
all_teams_agg <- all_teams_agg |> 
  inner_join(teams_divisions, by = c("team"="name"))

```

## Export Data
We'll export the data to flat files so they can be imported into Tableau.
```{r export dataframes}
write_excel_csv(all_teams, file = "day_night_by_team.csv")
write_excel_csv(all_teams_agg, file = "day_night_wins_by_team.csv")
```

